\chapter*{References}

[1] Preskill, J., \textit{Quantum Computing and the Entanglement Frontier}, arXiv:1203.5813, 2012. Available at: https://arxiv.org/pdf/1203.5813\\

\noindent[2] Arute, F., et al., \textit{Quantum Supremacy using a Programmable Superconducting Processor}, Nature, vol. 574, pp. 505–510, 2019. Available at: https://www.nature.com/articles/s41586-019-1666-5\\

\noindent[3] Dirac, P. A. M., \textit{The Principles of Quantum Mechanics}, 4th ed., Oxford University Press, 1958.\\

\noindent[4] Pednault, E., Maslov, D., Gunnels, J., Gambetta, J., \textit{IBM Quantum Computing Blog | On “quantum supremacy”}, Available at: https://www.ibm.com/quantum/blog/on-quantum-supremacy, Visited March 08, 2025.\\

\noindent [5] Busch, P., Heinonen, T., \& Lahti, P., \textit{Heisenberg's Uncertainty Principle}, arXiv:quant-ph/0609185v3, 2007. Available at: https://arxiv.org/abs/quant-ph/0609185 \\

\noindent[6] Artin, M., \textit{Algebra}, Pearson, 2010.\\

\noindent[7] Hoffman, K., \& Kunze, R., \textit{Linear Algebra}, Prentice-Hall, 1971.\\

\noindent[8] Axler, S., \textit{Linear Algebra Done Right}, Springer, 1997.\\

\noindent[9] Nielsen, M. A., \& Chuang, I. L., \textit{Quantum Computation and Quantum Information}, Cambridge University Press, 2010.\\

\noindent[10] Meyer, C. D., \textit{Matrix Analysis and Applied Linear Algebra}, SIAM, 2000.\\

\noindent[11] Sakurai, J. J., \& Napolitano, J., \textit{Modern Quantum Mechanics}, 3rd ed., Cambridge University Press, 2020.\\

\noindent[12] Arias-Castro, E., \textit{Principles of Statistical Analysis}, v1.0, University of California, San Diego, 2021.\\

\noindent[13] Hall, B. C., \textit{Quantum Theory for Mathematicians}, Springer, 2013.\\

\noindent[14] Rynne, B., and Youngson, M., \textit{Linear Functional Analysis}, Springer, 2008.\\

\noindent[15] Easttom, C., \textit{Quantum Computing Fundamentals}, Addison-Wesley, 2021. Available at: https://ptgmedia.pearsoncmg.com/images/9780136793816/samplepages/9780136793816\_Sample.pdf\\

\noindent[16] Griffiths, D. J., \textit{Introduction to Quantum Mechanics}, 2nd ed., Pearson Prentice Hall, 2005.\\

\noindent[17] Koprinkov, I. G., \textit{Spatial, Temporal and Coherent Superposition of Quantum States, a Reinterpretation of the Superposition Principle}, Journal of Modern Physics, Vol. 15, No. 2, pp. 183–197, 2024. Available at: https://www.scirp.org/journal/paperinformation.aspx?paperid=137875\\

\noindent[18] Marinescu, D. C., \& Marinescu, G. M., \textit{Classical and Quantum Information}, Academic Press, 2011.\\

\noindent[19] Greenberger, D., Hentschel, K., \& Weinert, F. (Eds.), \textit{Compendium of Quantum Physics: Concepts, Experiments, History and Philosophy}, Springer, 2009.\\

\noindent[20] Tokmakoff, A., \textit{Time-Dependent Quantum Mechanics and Spectroscopy}, University of Chicago, Lecture Notes.\\

\noindent[21] Feynman, R. P., Leighton, R. B., \& Sands, M., \textit{The Feynman Lectures on Physics, Vol. 3}, Addison-Wesley, 1965.\\

\noindent[22] Kalami Heris, M., \textit{Unveiling the Inner Product: The Key to Similarity in Math, Machine Learning, and Beyond}, Available at: https://kalami.medium.com/unveiling -the-inner-product-the-key-to-similarity-in-math-machine-learning-and-beyond-290b43cf34fd, Visited March 30, 2025.\\

\noindent[23] Peeters, K., \textit{Commutators and the Uncertainty Principle}, University of Durham, Available at: https://www.maths.dur.ac.uk/users/kasper.peeters/mathphys/uncertainty.html, Visited March 30, 2025.\\

\noindent[24] Maziero, J., \textit{Computing Partial Traces and Reduced Density Matrices}, arXiv:1601.07458, 2016. Available at: https://arxiv.org/pdf/1601.07458\\

\noindent[25] LaRose, R., \textit{Quantum States and Partial Trace}, QuIC Seminar 6, Michigan State University, 2018. Available at: https://www.ryanlarose.com/uploads/1/1/5/8/115879647/quic06-states-trace.pdf\\

\noindent[26] Wikipedia- Smite-Meister , \textit{Bloch sphere}, Wikipedia, The Free Encyclopedia. Available at: https://en.wikipedia.org/wiki/Bloch\_sphere, Visited April 2, 2025.\\

\noindent[27] Ladyman, J., Presnell, S., Short, A. J., \& Groisman, B., \textit{The connection between logical and thermodynamic irreversibility}, Studies in History and Philosophy of Modern Physics, Vol. 38, Issue 1, 2007. Available at: https://www.sciencedirect.com/science/article/pii/S1355219806000682\\

\noindent[28] Grover, L. K., \textit{A fast quantum mechanical algorithm for database search}, arXiv:quant-ph/9605043, 1996. Available at: https://arxiv.org/abs/quant-ph/9605043\\

\noindent[29] Yanofsky, N. S., \& Mannucci, M. A., \textit{Quantum Computing for Computer Scientists}, Cambridge University Press, 2008. Available at: https://quantumatlas.ir/wp-content/uploads/2025/01/Quantum-Computing-for-Computer
\noindent-Scientists.pdf\\

\noindent[30] IBM Quantum Learning, Fundamentals of Quantum Algorithms \textit{Grover's Algorithm}, Available at: https://learning.quantum.ibm.com/course/fundamentals-of-quantum-algorithms
\noindent/grovers-algorithm, Visited April 5, 2025.\\

\noindent[31] Youvan, D., \textit{Microsoft's Majorana 1: A Paradigm Shift Toward Scalable and Fault-Tolerant Quantum Computing}, 2025. Available at: https://doi.org/10.13140/RG.2.2.11176.89607\\

\noindent[32] Microsoft Quantum, \textit{Types of Qubits}, Microsoft Quantum Documentation. Available at: https://quantum.microsoft.com/en-us/insights/education/concepts/types-of-qubits, Visited April 6, 2025.\\

\noindent[33] Bedington, R., Arrazola, J. M., \& Ling, A., \textit{Progress in Satellite Quantum Key Distribution}, npj Quantum Information, vol. 3, article 30, 2017. Available at: https://doi.org/10.1038/s41534-017-0031-5\\

\noindent[34] Georgescu, I. M., Ashhab, S., \& Nori, F., \textit{Quantum Simulation}, Reviews of Modern Physics, vol. 86, no. 1, pp. 153–185, 2014. Available at: https://doi.org/10.1103/RevModPhys.86.153\\

\noindent[35] Alexeev, Y., Farag, M. H., Patti, T. L., Wolf, M. E., Ares, N., Aspuru-Guzik, A., Benjamin, S. C., Cai, Z., Chandani, Z., Fedele, F., Harrigan, N., Kim, J.-S., Kyoseva, E., Lietz, J. G., Lubowe, T., McCaskey, A., Melko, R. G., Nakaji, K., Peruzzo, A., Stanwyck, S., Tubman, N. M., Wang, H., \& Costa, T., \textit{Artificial Intelligence for Quantum Computing}, arXiv:2411.09131 [quant-ph], 2024. Available at: https://arxiv.org/abs/2411.09131\\


\noindent[36] Brandão, F. G. S. L., Christandl, M., Harrow, A. W., \& Walter, M., \textit{The Mathematics of Entanglement}, arXiv:1604.01790, 2016. Available at: https://arxiv.org/pdf/1604.01790\\

\noindent[37] Elbaset, M. A., Deraz, S. A., Soliman, M. A., \& Moussa, K. H., \textit{A 32-bit Quantum Encryption Algorithm Using Dynamic Pauli Gates}, 2023. Available at: https://www.researchgate.net/publication/372648350\_A\_32-bit\_Quantum\_Encryption\_
\noindent Algorithm\_Using\_Dynamic\_Pauli\_Gates\\

\noindent[38] Lambropoulos, P., \& Petrosyan, D., \textit{Fundamentals of Quantum Optics and Quantum Information}, Springer, 2007. Available at: https://www.researchgate.net/publication/253619080\_Fundamentals\_of\_Quantum\_Optics\_
\noindent and\_Quantum\_Information\\

\noindent[39] IBM Quantum Learning, \textit{Quantum Circuits}, Available at: https://learning.quantum.ibm.com/course/basics-of-quantum-information/quantum-circuits, Visited April 4, 2025.\\


